---
  title: "95-845: AA-MLP Project"
  author: Eileen Patten
  output:
    html_document
  fig_width: 7
  fig_height: 5
---

```{r, warning=FALSE, echo=FALSE, include=FALSE}
### Load helper packages ###
loadlibs = function(libs) {
  for(lib in libs) {
    class(lib)
    if(!do.call(require,as.list(lib))) {install.packages(lib)}
    do.call(require,as.list(lib))
  }
}
libs = c("tidyr","magrittr","purrr","dplyr","stringr","readr","data.table", "lubridate", "qwraps2", "randomForest", "pROC", "ggplot2", "knitr", "twang", "glmnet", "scales", "keras")
loadlibs(libs)


options(scipen=999)
```

The original dataset has 947 features and 131,076 observations. We classify the features broadly as features from previous history, current crime, from the person (demographics), guidelines of sentence, movements during incaceration, and features describing behavior after release.
```{r}
setwd("C:/Users/EileenPatten/Documents/MLPipeline/Project") 
load(file = "PCS_Recidivism_04_06_AlexC.Rda") #original dataset
```

###THIS SECTION OF CODE WAS CREATED BY ONE OF MY SYSTEMS TEAMMATES 
We selected 649 out of them, mainly deleting the movements of the inchoates while in jail and administrative variables from the "current crime".
```{r}
#Commented out because I build this into the synthetic data file
red <- df %>% select(2:5, 7, 14:16, 18,
                       126, 630, 631, 944:946, 680:683,#sex and race
                       171:174,  # offender age at time of offense, dates of offense and sentence
                       186, 191, # arrest date
                       21:23, #judge name DELETE
                       34, #PK of offenses
                       61, #completed or uncompleted crime
                       68,142, #type of incarceration (if any)
                       84,89:90, 636:638, 646:662, 739, #about the crime and it seriousness
                       678, 942, # OGS scores
                       117:125, 605, # punishment info
                       43, 193, 629, 738, # sentence start, year of sentence, county
                       130:133, 141, 150, # probation dates and indicators
                       52, #comment on the sentence
                       53, 97:99 , #conformity to guideline and reasons for being out of the guidelines
                       59, 102:106, #guidelines
                       155, 194,# sentencing guideline level (5 colors in matrix)
                       55, 113, #info of drug amount
                       56:58 , 164, #enhancements increasing OGS
                       286:327, 671:677, # prior charges
                       328:367, # juvenile record
                       368:407, #adult record
                       408:447, 697:729, # first charges
                       448:487, 623:628, # yr1 after release felonies
                       488:527, 617:622, # yr2 felonies
                       528:567, 689:692, 695:696, # yr3 felonies
                       569:572, 574:576, 585, 741, # Release date,escape flag, sex, and race prior
                       607, # Type of sanction given;
                       684:685, 744, 743, 693:694,  # years of follow up, rearrest
                       734,737,740,745,751,
                       753:ncol(df))

```

###THIS SECTION OF CODE WAS CREATED BY ONE OF MY SYSTEMS TEAMMATES 
In many cases if there were not prior charges, the feature AgeFirstArrest was missing. Then, we imputed it with the age of the offender, modifying 1400 cases. With this, we only have 11 missing cases.
```{r}
red<-red %>%
  mutate(AgeFirstArrest = ifelse(totalpriorcharges==0,floor(dofage),AgeFirstArrest) #imputing missing data on AgeFirstArrest
        )
```

##ADDED BY EILEEN -- A BIT MORE STRAIGHTFORWARD THAN THE WAY WE WERE DOING IT IN OUR SYSTEMS CODE
Because there are some dob entered incorrectly in the system, there are some offenders who have a date of offense age that is negative or less than 18 (907). Dropping these cases. 
```{r}
#table(as.integer(red$dofage))
#table(red$dofage<18)
red<- red %>%
  filter(is.na(dofage) | (dofage >= 18 & dofage <= 90)) 
         
```

###THIS SECTION OF CODE WAS CREATED BY ONE OF MY SYSTEMS TEAMMATES 
###I ADDED THE CEILING AT 90 BECAUSE I DON'T THINK THAT AGES ABOVE THAT ARE VALID
RAP sheets data on juvenile records is not accurate, so we put the minimum age of first arrest on 18. This modifies 21747 cases. 16661 of those cases were 15 or older. 41 were negative and 45 were zero. Since these individuals likely had some juvenile crimes, but this just isn't likely a full record, we think a base of 18 is reasonable. It is more likely that there are now some over 18 for age of first crime that should be set to 18 but had missing juvenile records. Since we only have a date for the first crime in their record, there is nothing we can do about this. 
```{r}
red<-red %>%
  mutate(AgeFirstArrest18 = ifelse(AgeFirstArrest<18,18,
                                   ifelse(AgeFirstArrest>90,90,
                                          AgeFirstArrest))) 

#table(red$AgeFirstArrest18)                                  
```

###THIS SECTION OF CODE WAS CREATED BY ONE OF MY SYSTEMS TEAMMATES BUT EDITED BY EP
Additionally, we created new variables related to the length of criminal career, total convictions and number of crimes per year.
```{r}
red <- red %>%
  mutate(age_difference = dofage - AgeFirstArrest18, #lenght of criminal career (as an adult)
         tot_conv = rowSums(select(., contains("PRSCONV"))), #total convictions
         age_diff_sq = age_difference^2, #criminal career squared
         age_diff_aux = ifelse(age_difference<1,1,age_difference),
         rate_offense = totalpriorcharges_adult/age_diff_aux  #number of offenses per year of criminal careeer
         ) 

```

###THIS SECTION OF CODE WAS CREATED BY ONE OF MY SYSTEMS TEAMMATES BUT EP ADDED ALL THE DRUG AND WEAPON VARIABLES
Transforming characters into classes (factor)
```{r}
#drug -- MAKE NA EQUAL TO 0 BECAUSE IF THERE IS NO RECORD IT MEANS THERE WAS NO DRuG INVOLVED
red$drug[is.na(red$drug)] <- "None"


red<- red %>%
  transform(OFF_SEX = as.factor(OFF_SEX),
            drug = as.factor(drug), 
            POSSESSED_USED = as.factor(POSSESSED_USED), 
            ENHANCEMENT_WEAPONS = as.factor(ENHANCEMENT_WEAPONS)
        )
```

###THIS SECTION OF CODE WAS CREATED BY ONE OF MY SYSTEMS TEAMMATES 
###THERE ARE FEWER OF THESE SINCE I DROPPED THE NEGATIVE OFFENSE AGES AND WE CAPPED AGE AT FIRST AT 18 BUT SINCE THERE ARE PEOPLE THAT ARE CLEARLY ERRONEOUSLY OLD AT DOFAGE AND AGEFIRSTARREST THERE ARE STILL SOME NEGATIVES
Also, we filtered cases with age difference negative (we keep cases until -1, to avoid problems of age coding integer versus continuous) and 3 cases with age difference greater than 75. We end with 129727 cases
```{r}
#table(red$dofage>red$AgeFirstArrest18)
#table(as.integer(red$dofage))
#table(as.integer(red$AgeFirstArrest18))

red<- red %>%
  filter(is.na(age_difference) | (age_difference >= -1 & age_difference <= 75) #to avoid problems of continuous vs integer ages
         ) 

nrow(red)
```

####NEW EP DATA CLEANING/CHECKING -- FROM THIS POINT FORWARD THIS IS ALL ORIGINAL CODE
Since this ML project is about the type of first arrest, I want to make sure the first arrest data looks okay. 
```{r}
#50655 do not recidivate
table(red$totalchargesfirstarrest, exclude = NULL)

#Examine the variables -- determine which are unique and which are different versions of the same
first <- grep("^firstburglaryt$", colnames(red)) #230
#last <- grep("^firstPublicOrdert$", colnames(red)) #245
#last <- grep("^firstkidnappingt$", colnames(red)) #249
#last <- grep("^firstviolent_middlet$", colnames(red)) #267
last <- grep("^firstabscondt$", colnames(red)) #247
red <- red %>%
  mutate(first_sum = Reduce("+",.[first:last])) #total charges first arrest should match var of same name


table(red$totalchargesfirstarrest==red$first_sum)
#table(red$firstmurdert[red$firstPersonalFt==0 & red$firstPersonalMt==0])
#NOTE: MURDER IS INCLUDED IN firstPErsonalFt BUT SHOULD BE CODED INTO ITS OWN CATEGORY BECAUSE IT IS WORSE THAN AN F
#Need to take murders out of firstPErsonalFt
red$firstPersonalFt <- red$firstPersonalFt - red$firstmurdert
red <- red %>%
  mutate(first_sum = Reduce("+",.[first:last]) + firstmurdert) #total charges first arrest should match var of same name
table(red$totalchargesfirstarrest==red$first_sum)

#For the future crime (variables with "first" in the name), I know the type to crime (burglary, kidnapping, etc) but not the grade (Non-graded, Misdemeanor, Felony, Murder). In order to determine the grade, I am examining the civil code, and where it is not clear, I am looking at the matching variables on crime type for the current crime (for which I do have grade information) to determine what grade each category should be. In most cases this assignment was based on 97% or more (one was 80%). Since people often commit more than one crime in each arrest, I will take the most serious crime. 

#This is to determine how to code the "first" variables which match up to these variables. My assumption is that beaucause these are determined by law if the current burglaries are all felonies (or almost all) then felonies in the next crime variable will also be felonies
#find variable indices
first <- grep("^burglary$", colnames(red))
last <- grep("^publicOrder$", colnames(red))


#determining what these crimes are classified as, on average
for(index in first:last){
  print(colnames(red)[index])
  print(prop.table(table(red[[index]], red$grade),1))
  print('')
  print('')
}

#NOTE: No way to figure out domestic charges grade so I am dropping those from the data
red$totalchargesfirstarrest <- red$totalchargesfirstarrest - red$firstdomestict


red$nongrade <- ifelse((red$firstDUIt>0 | red$firstconspiracyt>0),1,0)
#I checked these all like this
#table(red$nongrade[red$firstDUIt==0 & red$firstconspiracyt==0])

red$misdemeanor <- ifelse((red$firstpropertyMt > 0 |      #100% are misdemeanors for current crime (also labeled M)
                          red$firstPersonalMt > 0 |       #100%
                          red$firstSexOMt > 0 |           #100%
                          red$firstDrugMt > 0 |           #100%
                          red$firstOtherTraffict > 0 |    #100%
                          red$firstOtherweaponst > 0 |    #99%
                          red$firstPublicAdmt > 0 |       #80%
                          red$firstPublicOrdert > 0 |     #97%
                          red$firstabscondt > 0),         #100%
                          1,0)         

red$felony <- ifelse((red$firstburglaryt > 0 |      #100% are felonies for current crime (also some are labeled F)
                      red$firstpropertyFt > 0 |     #100%
                      red$firstPersonalFt > 0 |     #100%
                      red$firstSexOFt > 0 |         #98%
                      red$firstDrugFt > 0 |         #99.8%
                      red$firstFirearmst > 0),      #78%
                      1,0)                         

red$murder <- ifelse((red$firstmurdert > 0), 1, 0)

#Now I need to select the most serious charge for each person
#4=murder; 3=felony; 2=misdemeanor; 1=nongraded (DUI and conspiracy); 0=no recidivism in 3 years
red$toprearrest <- ifelse(red$murder==1, 4,
                          ifelse(red$felony==1, 3,
                                 ifelse(red$misdemeanor==1, 2,
                                        ifelse(red$nongrade==1, 1, 0))))
red$toprearrest <- as.factor(red$toprearrest)
table(red$toprearrest)
#Checking -- this looks good -- # with NO arrest in totalchargesfirstarrest == # with 0 in new variable
#table(red$totalchargesfirstarrest>0)
#table(red$toprearrest)
#table(red$toprearrest, red$murder) -- and all murders are in category 4
```

```{r}
#Some of the PRSCONV variables are the same crime but have separate variables because they are coming from different editions of the sentencing guidelines. This does not have any practical meaning data-wise so I am collapsing them. 

#Example of variable and label
#PRSCONV_9	Number of prior aggravated assault (SBI) convictions
#PRSCONV_49	Number of prior aggravated assault (SBI) convictions
red$PRSCONV1_41  <- red$PRSCONV_1  + red$PRSCONV_41
red$PRSCONV2_42  <- red$PRSCONV_2  + red$PRSCONV_42
red$PRSCONV3_43  <- red$PRSCONV_3  + red$PRSCONV_43
red$PRSCONV4_44  <- red$PRSCONV_4  + red$PRSCONV_44
red$PRSCONV5_45  <- red$PRSCONV_5  + red$PRSCONV_45
red$PRSCONV8_48  <- red$PRSCONV_8  + red$PRSCONV_48
red$PRSCONV9_49  <- red$PRSCONV_9  + red$PRSCONV_49
red$PRSCONV10_59 <- red$PRSCONV_10 + red$PRSCONV_59
red$PRSCONV11_50 <- red$PRSCONV_11 + red$PRSCONV_50
red$PRSCONV12_58 <- red$PRSCONV_12 + red$PRSCONV_58
red$PRSCONV13_61 <- red$PRSCONV_13 + red$PRSCONV_61
red$PRSCONV18_62 <- red$PRSCONV_18 + red$PRSCONV_62
red$PRSCONV19_51 <- red$PRSCONV_19 + red$PRSCONV_51
red$PRSCONV20_53 <- red$PRSCONV_20 + red$PRSCONV_53
red$PRSCONV21_63 <- red$PRSCONV_21 + red$PRSCONV_63
red$PRSCONV22_65 <- red$PRSCONV_22 + red$PRSCONV_65
red$PRSCONV23_64 <- red$PRSCONV_23 + red$PRSCONV_64
red$PRSCONV24_66 <- red$PRSCONV_24 + red$PRSCONV_66
red$PRSCONV25_67 <- red$PRSCONV_25 + red$PRSCONV_67
red$PRSCONV26_68 <- red$PRSCONV_26 + red$PRSCONV_68
red$PRSCONV28_71 <- red$PRSCONV_28 + red$PRSCONV_71
red$PRSCONV29_72 <- red$PRSCONV_29 + red$PRSCONV_72
red$PRSCONV30_74 <- red$PRSCONV_30 + red$PRSCONV_74
red$PRSCONV31_75 <- red$PRSCONV_31 + red$PRSCONV_75
red$PRSCONV32_78 <- red$PRSCONV_32 + red$PRSCONV_78
red$PRSCONV33_79 <- red$PRSCONV_33 + red$PRSCONV_79
red$PRSCONV34_80 <- red$PRSCONV_34 + red$PRSCONV_80
red$PRSCONV35_81 <- red$PRSCONV_35 + red$PRSCONV_81
red$PRSCONV36_82 <- red$PRSCONV_36 + red$PRSCONV_82
red$PRSCONV37_83 <- red$PRSCONV_37 + red$PRSCONV_83
red$PRSCONV38_84 <- red$PRSCONV_38 + red$PRSCONV_84
red$PRSCONV39_76 <- red$PRSCONV_39 + red$PRSCONV_76
red$PRSCONV40_89 <- red$PRSCONV_40 + red$PRSCONV_89


#PRSCONV_7 and PRSCONV_46 are not labeled in our data dictionary, so I am dropping them
drop <- c("PRSCONV_9","PRSCONV_49","red$PRSCONV_1", "red$PRSCONV_41", "red$PRSCONV_2",  "red$PRSCONV_42", "red$PRSCONV_3",  "red$PRSCONV_43", "red$PRSCONV_4", "red$PRSCONV_44", "red$PRSCONV_5",  "red$PRSCONV_45", "red$PRSCONV_8",  "red$PRSCONV_48",  "red$PRSCONV_9",  "red$PRSCONV_49", "red$PRSCONV_10", "red$PRSCONV_59", "red$PRSCONV_11", "red$PRSCONV_50", "red$PRSCONV_12", "red$PRSCONV_58", "red$PRSCONV_13", "red$PRSCONV_61",  "red$PRSCONV_18", "red$PRSCONV_62", "red$PRSCONV_19", "red$PRSCONV_51", "red$PRSCONV_20", "red$PRSCONV_53", "red$PRSCONV_21", "red$PRSCONV_63", "red$PRSCONV_22", "red$PRSCONV_65",  "red$PRSCONV_23", "red$PRSCONV_64", "red$PRSCONV_24", "red$PRSCONV_66", "red$PRSCONV_25", "red$PRSCONV_67", "red$PRSCONV_26", "red$PRSCONV_68", "red$PRSCONV_28", "red$PRSCONV_71",  "red$PRSCONV_29", "red$PRSCONV_72", "red$PRSCONV_30", "red$PRSCONV_74", "red$PRSCONV_31", "red$PRSCONV_75", "red$PRSCONV_32", "red$PRSCONV_78", "red$PRSCONV_33", "red$PRSCONV_79",  "red$PRSCONV_34", "red$PRSCONV_80", "red$PRSCONV_35", "red$PRSCONV_81", "red$PRSCONV_36", "red$PRSCONV_82", "red$PRSCONV_37", "red$PRSCONV_83", "red$PRSCONV_38", "red$PRSCONV_84",  "red$PRSCONV_39", "red$PRSCONV_76", "red$PRSCONV_40", "red$PRSCONV_89", "PRSCONV_46","PRSCONV_7")
red = red[,!(names(red) %in% drop)]

###SELECTING THE FEATURES I WANT IN MY MODEL

###subset data to use in the models
#str(red, list.len=ncol(red))
#MANY of these variables are future-looking and inappropriate to include in the model or are factors the judge/judicial system would not have access to for one reason or another

#Make same change to murder as in data prep for the outcome variable
red$tPersonalF_adult <- red$tPersonalF_adult - red$tmurder_adult


model_data_selections<- c("toprearrest", "PRSCONV1_41", "PRSCONV2_42", "PRSCONV3_43", "PRSCONV4_44", "PRSCONV5_45", "PRSCONV_6", "PRSCONV8_48", "PRSCONV9_49", "PRSCONV10_59", "PRSCONV11_50", "PRSCONV12_58", "PRSCONV13_61", "PRSCONV_14", "PRSCONV_15", "PRSCONV_16", "PRSCONV_17", "PRSCONV18_62", "PRSCONV19_51", "PRSCONV20_53", "PRSCONV21_63", "PRSCONV22_65", "PRSCONV23_64", "PRSCONV24_66", "PRSCONV25_67", "PRSCONV26_68", "PRSCONV_27", "PRSCONV28_71", "PRSCONV29_72", "PRSCONV30_74", "PRSCONV31_75", "PRSCONV32_78", "PRSCONV33_79", "PRSCONV34_80", "PRSCONV35_81", "PRSCONV36_82", "PRSCONV37_83", "PRSCONV38_84", "PRSCONV39_76", "PRSCONV40_89", "PRSCONV_47",  "PRSCONV_52", "PRSCONV_54", "PRSCONV_55", "PRSCONV_56", "PRSCONV_57",  "PRSCONV_60", "PRSCONV_61",  "PRSCONV_69", "PRSCONV_70", "PRSCONV_73",  "PRSCONV_77", "PRSCONV_85", "PRSCONV_86", "PRSCONV_87", "PRSCONV_88", "tburglary_adult", "tpropertyF_adult", "tpropertyM_adult", "tPersonalM_adult", "tPersonalF_adult", "tSexOF_adult", "tSexOM_adult", "tDrugF_adult", "tDrugM_adult", "tDUI_adult", "tOtherTraffic_adult", "tFirearms_adult", "tOtherweapons_adult", "tdomestic_adult", "tabscond_adult", "tconspiracy_adult", "tPublicOrder_adult", "tPublicAdm_adult", "tmurder_adult", "AgeFirstArrest18", "rate_offense", "drug", "ENHANCEMENT_WEAPONS", "POSSESSED_USED", "race_new", "age_new", "dofage", "sex_new")
model_data <- red[model_data_selections]

###CHECKING ALL THE VARIABLES THAT REMAIN
#Drop any crime variables that have fewer than 50 people who have ever committed this crime in our data
#To find these cases
# for(i in 2:ncol(model_data)) {
#   true <- (model_data[i]>0)
#   numtrue <- sum(true)
#   print(colnames(model_data)[i])
#   print(numtrue)
# }

drop <- c("PRSCONV_6", "PRSCONV8_48", "PRSCONV10_59", "PRSCONV12_58", "PRSCONV_27", "PRSCONV30_74",
          "PRSCONV31_75", "PRSCONV33_79", "PRSCONV35_81",  "PRSCONV37_83", "PRSCONV_52",  "PRSCONV_54",
          "PRSCONV_55", "PRSCONV_56", "PRSCONV_57",  "PRSCONV_60", "PRSCONV_61", "PRSCONV_69",
          "PRSCONV_70", "PRSCONV_73", "PRSCONV_85")
model_data = model_data[,!(names(model_data) %in% drop)]

#table(model_data$AgeFirstArrest18)
#table(model_data$rate_offense)
#table(red$totalpriorcharges_adult)

#table(model_data$drug)
#Collapsing some levels of factors that are sparsely populated into "Other" categories
#Anything with less than 50 cases in drug
model_data$drug[model_data$drug=="Designer Drugs" | model_data$drug=="Steroids" | model_data$drug=="Ectasy"] <- "Other Drug Offenses"
model_data$drug <- factor(model_data$drug) #drops factors with 0 cases

# levels(model_data$ENHANCEMENT_WEAPONS)
#(model_data$ENHANCEMENT_WEAPONS)
#Collapsing the gun categories -- in this case I am going to leave blunt instrument even though it has 37 observations because there's not an obvious category to wrap it in with
levels(model_data$ENHANCEMENT_WEAPONS) <- c(levels(model_data$ENHANCEMENT_WEAPONS),"Firearm", "No Weapon", "Other/Unknown Weapon")
model_data$ENHANCEMENT_WEAPONS[model_data$ENHANCEMENT_WEAPONS=="HandGun/Pistol/Revolver" |
                               model_data$ENHANCEMENT_WEAPONS=="Long Gun/Rifle/Shotgun" |
                               model_data$ENHANCEMENT_WEAPONS=="Gun/Other"] <- "Firearm"
model_data$ENHANCEMENT_WEAPONS[model_data$ENHANCEMENT_WEAPONS==""] <- "No Weapon"
model_data$ENHANCEMENT_WEAPONS[model_data$ENHANCEMENT_WEAPONS=="Other Weapon" |
                               model_data$ENHANCEMENT_WEAPONS=="Unknown Weapon"] <- "Other/Unknown Weapon"
model_data$ENHANCEMENT_WEAPONS <- factor(model_data$ENHANCEMENT_WEAPONS) #drops factors with 0 cases

#table(model_data$POSSESSED_USED)

#table(model_data$race_new)
#Note: I am not trying to impute some of these, like race, because this is already an imputed variable
model_data$race_new[model_data$race_new=="Unknown"] <- "Other"
model_data$race_new <- factor(model_data$race_new) #drops factors with 0 cases

#table(model_data$age_new)
model_data$age_new <- factor(model_data$age_new) #drops factors with 0 cases

#table(as.integer(model_data$dofage))

#table(model_data$sex_new)
model_data$sex_new <- factor(model_data$sex_new) #drops factors with 0 cases

#table(model_data$tot_conv)


save(red, file = "FINAL_red_ML.Rda")


```